Supplementary material for  \nBayesian multi-proxy reconstruction of the early Eocene latitudinal temperature gradient

### 2023-01-23 manuscript/manuscript.Rmd
SM and results

## Parameter estimation
We estimated the posterior distributions of the model parameters using a Markov chain Monte Carlo (MCMC) algorithm, written in R. Specifically, we sampled the unknown parameters $A$, $K$, $M$ and $B$ with Metropolis-Hastings, and used Gibbs sampling to estimate all other unknown parameters. Posterior inference on the modern gradient is based on four chains with 60,000 iterations each, 10,000 of which were discarded as burn-in, and keeping every 10th iteration, resulting in a total of 20,000 iterations. The modern temperature gradient with the Eocene sampling distribution was estimated in one chain with 25,000 iterations for each of the 100 temperature samples. 5,000 iterations each were discarded as burn-in, and every 10th iteration was kept, resulting in a total of 200,000 iterations across all 100 model runs.
For the Eocene model, we ran four chains with 600,000 iterations each, discarding 100,000 as burn-in and keeping every 100th iteration, as the hierarchical model structure results in higher autocorrelation of the chains. The Eocene posterior inference is thus based on a total of 20,000 iterations with low autocorrelation (effective multivariate sample size for $A$, $K$, $M$ and $B$ is > 18,000).

@@

# Results

## Model validation
<!-- - Show that the 4-parameter logistic function can adequatly describe the modern temperature gradient
- Show that the modern gradient can be reproduced with Eocene sampling distribution
-->
[Fig 3 - Modern gradient]
The average modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig. 3a). Degrading the amount and distribution of data points to match the early Eocene sampling distribution ($n = 34$) still results in a close match of the modelled gradient with the empirical gradient (Fig 3b). A noticable, but minor offset is apparent in high latitudes, where sampling is most incomplete (Fig 3b).

## EECO reconstruction
<!-- - Show Eocene gradient
- Global t : use raster package for calculation of areal weights
- report lat gradient
-->
```{r results block - for in-line code, echo = FALSE}
setwd('..') # move up to parent folder of /manuscript

# temperature gradients
tdiff_lat0 <- readRDS("results/eeco/eocene_modern_difference_equator.rds")
tdiff_lat90 <- readRDS("results/eeco/eocene_modern_difference_poles.rds")
eocene_gradient <- readRDS("results/eeco/eocene_overall_gradient.rds")
modern_gradient <- readRDS("results/modern/modern_overall_gradient.rds")

@@

# global average temperatures
motemp_global <- readRDS("results/modern/global_mean_modern.rds")
eotemp_global <- readRDS("results/eeco/global_mean_eeco.rds")

```

@@

The high variability of early Eocene palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, results in substantial uncertainties in the modeled temperature gradient. This is reflected in the residual standard deviation ($\sigma$) of the early Eocene gradient, `r eeco_sdy$"50%"` (95% CI: `r eeco_sdy$"2.5%"` - `r eeco_sdy$"97.5%"`) degC, more than twice as high than  $\sigma$ of the modern gradient, resampled at early Eocene latitudes, `r modern_eeco_sample_sdy$"50%"` (95% CI: `r modern_eeco_sample_sdy$"2.5%"` - `r modern_eeco_sample_sdy$"97.5%"`) degC. The inclusion of ecological proxy data only slightly reduces the overall model standard deviation, but helps constrain polar temperature estimates (Supplementary materials).

>>>>

## Parameter estimation
We estimated the posterior distributions of the model parameters using a Markov chain Monte Carlo (MCMC) algorithm, written in R. Specifically, we sampled the unknown parameters $A$, $K$, $M$ and $B$ with Metropolis-Hastings, and used Gibbs sampling to estimate all other unknown parameters. Posterior inference on the modern gradient is based on four chains with 60,000 iterations each, 10,000 of which were discarded as burn-in, and keeping every 10th iteration, resulting in a total of 20,000 iterations. The modern temperature gradient with the Eocene sampling distribution was estimated in one chain with 25,000 iterations for each of the 100 temperature samples. 5,000 iterations each were discarded as burn-in, and every 10th iteration was kept, resulting in a total of 200,000 iterations across all 100 model runs.
For the Eocene model, we ran four chains with 600,000 iterations each, discarding 100,000 as burn-in and keeping every 100th iteration, as the hierarchical model structure results in higher autocorrelation of the chains. The Eocene posterior inference is thus based on a total of 20,000 iterations with low autocorrelation (effective multivariate sample size for $A$, $K$, $M$ and $B$ is > 18,000). Trace plots were used to assess the convergence and mixing of the MCMC chains (Supplementary Materials S1).

@@

# Results
```{r results block - for in-line code, echo = FALSE}
setwd('..') # move up to parent folder of /manuscript

# temperature gradients
tdiff_lat0 <- readRDS("results/eeco/eocene_modern_difference_equator.rds")
tdiff_lat90 <- readRDS("results/eeco/eocene_modern_difference_poles.rds")
eocene_gradient <- readRDS("results/eeco/eocene_overall_gradient.rds")
modern_gradient <- readRDS("results/modern/modern_overall_gradient.rds")
modern_eocene_gradient <- readRDS("results/modern/modern_eocene_sampling_overall_gradient.rds")
modern_eocene_gradient_diff <- readRDS("results/modern/modern_eocene_sampling_overall_gradient_diff.rds")

@@

# global average temperatures
motemp_global <- readRDS("results/modern/global_mean_modern.rds")
eotemp_global <- readRDS("results/eeco/global_mean_eeco.rds")
moeotemp_global <- readRDS("results/eeco/global_mean_modern_eocene_sampling_eeco.rds")
moeotemp_global_diff <- readRDS("results/eeco/global_mean_modern_eocene_sampling_difference.rds")

```
## Model validation
<!-- - Show that the 4-parameter logistic function can adequatly describe the modern temperature gradient
- Show that the modern gradient can be reproduced with Eocene sampling distribution
-->
[Fig 3 - Modern gradient]
The average modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig. 3a). Reducing the amount and the spatial coverage of data points to match the early Eocene sampling distribution ($n = 34$) still results in a close match of the modeled gradient with the empirical gradient (Fig. 3b). The 95% credible interval of reconstructed gradients (Fig. 3b) is noticeably narrower than the spread of empirical temperature values (Fig. 3a). The modern, latitudinal gradient, modeled with all data, is `r modern_gradient$"50%"` degC, and has a median of `r modern_eocene_gradient$"50%"`  (95% CI: `r modern_eocene_gradient$"2.5%"` - `r modern_eocene_gradient$"97.5%"`) degC when modelled with only the early Eocene sampling locations. A random sampling of modern temperatures at latitudes corresponding to the sampled Eocene palaeolatitudes would thus allow the reconstruction of the modern latitudinal temperature gradient with good accuracy, with expected deviations from the actual gradient as low as `r modern_eocene_gradient_diff$"2.5%"` to `r modern_eocene_gradient_diff$"97.5%"` degC (95 % CI).

The modeled, latitudinal temperature gradient facilitates accurate reconstruction of global average temperatures. The modern, global mean sea surface temperature estimate is `r motemp_global$"50%"` degC when using the full, modern data set. When reduced to the early Eocene sampling distribution, the modern global mean sea surface temperature is estimated at `r moeotemp_global$"50%"` (95% CI: `r moeotemp_global$"2.5%"` - `r moeotemp_global$"97.5%"`) degC. The deviations from the full modern estimate are again very low, with a 95 % credible interval of `r moeotemp_global_diff$"2.5%"` to `r moeotemp_global_diff$"97.5%"` degC.

## EECO reconstruction
<!-- - Show Eocene gradient
- Global t : use raster package for calculation of areal weights
- report lat gradient
-->

@@

The high variability of early Eocene palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, results in substantial uncertainties in the modeled temperature gradient. This is reflected in the residual standard deviation ($\sigma$) of the early Eocene gradient, `r eeco_sdy$"50%"` (95% CI: `r eeco_sdy$"2.5%"` - `r eeco_sdy$"97.5%"`) degC, more than twice as high than  $\sigma$ of the modern gradient, resampled at early Eocene latitudes, `r modern_eeco_sample_sdy$"50%"` (95% CI: `r modern_eeco_sample_sdy$"2.5%"` - `r modern_eeco_sample_sdy$"97.5%"`) degC. The inclusion of ecological proxy data only slightly reduces the overall model standard deviation, but helps constrain polar temperature estimates (Supplementary materials).  ### Continue here.

### 2023-01-23 manuscript/supplementary_material.Rmd
update SM
SM and results

---
## Hierarchical, Bayesian model for latitudinal temperature gradients

**Model structure (Fig 2).** We model the mean temperature ($\mu$) at location $j$ as a function of absolute latitude ($abs(l)$) with a logistic regression (`growth curve` or `Richard's curve`) of the form:

$$\mu_{j}  \sim N(\nu_j, \sigma), $$

$$\nu_{j} = A~+\frac{K-A}{e^{B(abs(l_j)-M)}}~, ~~~~~ j = 1,...,n,  $$
where $A$ and $K$ denote the lower and upper asymptote, respectively, $M$ specifies the latitude of maximal growth, the most quickly with latitude, $B$ denotes the growth rate (Fig. 2), $\sigma$ denotes the residual standard deviation, and $n$ denotes the number of locations.

We infer $\mu_j$ from $m$ individual temperature observations $t_{i = 1,...m}$, derived from geochemical data, at location $j$ as

$$t_{i,j}  \sim N(\mu_j, \sigma_j), ~~~~~ i = 1,...,m,  $$
where $m$ is the number of observations at each location, and $\sigma_j$ is the estimated standard deviation of the temperatures at location $j$.

Similarly, $\mu_j$ is inferred for locations with ecological proxies from the associated normal temperature distributions with a given mean and standard deviation, $t_{\mu,j}$ and $t_{\sigma,j}$, as

$$t_{\mu,j}  \sim N(\mu_j,t_{\sigma,j}).  $$

This structure implies that $\mu_j$ is not fixed at the mean proxy temperature at location $j$, <!--of the temperatures $t_{i=1,...,m,~j}$ of location $j$-->, but is drawn towards the overall logistic regression curve, i.e. towards $\nu_j$. The pull towards $\nu_j$ tends to be high when $m$ is low, when the observations $t_{i=1,...,m,~j}$ are scattered, i.e. $\sigma_j$ is high, and/or when the overall standard deviation $\sigma$ is low. In practice, this has the desirable consequence that locations with few observations and large temperature differences between observations have less influence on the overall regression than well-sampled locations with consistent temperature observations.

**Priors (Fig2x).** In the Bayesian framework, priors need to be placed on the unknown parameters of a model. We placed weakly informative, conjugate inverse-gamma priors on $\sigma$ and $\sigma_{j=1,..n}$:
$$\sigma  \sim \sqrt{Inv-Gamma(\alpha+\frac{n}{2},\beta + 0.5\times(\mu_j-\nu_j))}, ~~~~~ j = 1,...,n,  $$
$$\sigma_j  \sim \sqrt{Inv-Gamma(\alpha+\frac{m}{2},\beta + 0.5\times(t_{i,j}-\mu_j))}, ~~~~~ i = 1,...,m,~~~~~j = 1,...,n. $$
We set $\alpha = \beta = 1$, allowing these priors to be quickly overwhelmed by the data as $n$ and $m$ increase, as we have little *a priori* knowledge of these parameters.

In contrast, we put informative priors on the regression coefficients $A$, $K$, $M$ and $B$, based on physical principles, and vaguely based on the modern climate system:

**A.** Predicted seawater surface temperatures are not allowed to be $<< -2^\circ C$, the freezing point of sea water. The highest prior density of $A$ is placed around $0^\circ C$, and it slowly tapers off towards higher temperatures. This shape is achieved by placing a skew-normal prior on the lower asymptote, specified as
$$A \sim SN(\xi=-3.0,\omega=12,\alpha_{SN}=30),$$
where $\xi$, $\omega$, and $\alpha_{SN}$ are the location, scale and shape parameters.

**K.** Input of solar energy decreases from the tropics to the poles. Hence, the latitudinal temperature gradient is broadly negative, i.e. temperature decreases with absolute latitude. This is achieved by setting $K \ge A$. The prior on the upper asymptote $K$ is a truncated normal distribution with the mean set to $K$ of the modern SST gradient, with a broad standard deviation:
$$K \sim TN(\mu_{TN} = 28, \sigma_{TN} = 10, \alpha_{TN} = A, \beta_{TN}= \infty)$$
The distribution is truncated to the left at $\alpha_{TN}=A$, but not truncated to the right ($\beta_{TN}$).

**M.** The steepness of the gradient is presumed to be highest in mid-latitudes; this is expressed with a normal prior on $M$ with the mean set to $M$ of the modern SST gradient:
$$M \sim N(42,10)$$
**B.** The steepness or growth rate $B$ of the gradient is constrained to be $\ge0$ and to not be exceedingly high, as oceanic and atmospheric heat transfer is bound to prevent very abrupt SST changes across latitudes. A gamma-distributed prior of the form
$$B \sim Gamma(\alpha_G=4.3,\beta_G=30)$$
was placed on $B$. The shape and rate parameters $\alpha_G$ and $\beta_G$ were chosen such that the highest prior density is at $B$ of the modern SST gradient, $0.11$.

# References

>>>>

---
```{r, setup, include=FALSE}
#knitr::opts_knit$set(root.dir = '..')
```

## S1. Convergence checks
To ensure the convergence and mixing of the chains, we inspected traceplots of all the unknown parameters of the model.

```{r out.width='100%', echo = FALSE, fig.cap = "Fig S1. Traceplots of a subset of the unknown model parameters. The four colours correspond to the four independent model runs. a) Traceplot of A, K-A, M, B and $\\sigma$; b) Traceplot of five selected $\\mu_j$; c) Traceplot of five selected $\\sigma_j$."}
setwd('..') # move up to parent folder of /manuscript
knitr::include_graphics("D://Github/PalaeoClimateGradient/figures/SM/FigS1-Traceplot.jpg", error = FALSE)
```

## S2. Gradient with just the geochemical proxy data

```{r out.width='100%', echo = FALSE, fig.cap = "Fig S2. Latitudinal temperature gradient with just the geochemical proxies (orange), showing the median (line) and 95% credibal interval (shading). Symbols with vertical lines show the median and 95% credible intervals of $\\mu$. The grey line and shading in the background show the latitudinal temperature gradient with the geochemical and ecological proxy data."}
setwd('..') # move up to parent folder of /manuscript
knitr::include_graphics("D://Github/PalaeoClimateGradient/figures/SM/FigS2_geochem_eeco.png", error = FALSE)
```

## S3. Seperate gradients on the northen and southern hemisphere

```{r out.width='100%', echo = FALSE, fig.cap = "Fig S3. Latitudinal temperature gradient in seperate hemispheres, showing the medians (lines) and 95% credibal intervals (shadings) in the southern (red) and northern hemisphere (blue). Symbols with vertical lines show the median and 95% credible intervals of $\\mu$. Turquoise symbols in the northern hemisphere highlight the ecological proxy data. The grey line and shading in the background show the latitudinal temperature gradient with the data from both hemispheres combined, plotted in both hemispheres."}
setwd('..') # move up to parent folder of /manuscript
knitr::include_graphics("D://Github/PalaeoClimateGradient/figures/SM/FigS3_north_south_eeco.png", error = FALSE)
```
